---
title: "How to implement BigQuery dataset into r"
author: Monika B
format: html
editor: visual
---

## What is BigQuery

BigQuery is Google's serverless data warehouse, with built-in features like machine learning and business intelligence to manage and analyze big enterprise data. It can be accessed as a limited, free tier service or subscribed to. It is very intuitive to use.

## Selecting a public database

There are multiple databases publicly available on Google Cloud within the free tier, my interest was on the ebi_chembl dataset. This is a manually curated database of small molecules used in drug discovery, including information about existing patented drugs. The
[Schema](http://ftp.ebi.ac.uk/pub/databases/chembl/ChEMBLdb/releases/chembl_23/chembl_23_schema.png) and the [Documentation](http://ftp.ebi.ac.uk/pub/databases/chembl/ChEMBLdb/releases/chembl_23/schema_documentation.html) of the database are available online.

## Writing SQL script to query the database

A SQL script was written to filter the public database for the available mesalamine products. SQL statements, like SELECT, JOIN and WHERE were used to join multiple databases on *molecule registration number* and *product id* foreign keys, after consulting the schema.

The results table and SQL files were saved into the BigQuery Project and Database.

## Load the data from BigQuery to r

First the necessary libraries are (downloaded if necessary and) imported into RStudio.

```{r}
#| output: false
library(bigrquery)
library(tidyverse)
```

Next, GCP BigQuery needs to authenticate this local RStudio project. For this, run the command bq_auth and log into the Google Account used for BigQuery and allow access (note: output is not shown)

```{r}
#| output: false
bq_auth()
```

The id of the project from GCP can be found on the landing page of the website. Also add the SQL query saved to the same project.

```{r}
project_id = 'learn-bigquery040625'
```

```{r}
my_sql_query <- "
  SELECT m.chembl_id, m.pref_name, d.mechanism_of_action, p.trade_name, p.route, p.dosage_form, p.applicant_full_name
FROM `patents-public-data.ebi_chembl.molecule_dictionary` m
JOIN `patents-public-data.ebi_chembl.drug_mechanism` d
ON m.molregno = d.molregno
JOIN `patents-public-data.ebi_chembl.formulations` f
ON d.molregno = f.molregno
JOIN `patents-public-data.ebi_chembl.products` p
ON f.product_id = p.product_id
WHERE m.chembl_id = 'CHEMBL704'
"
```

This next code chunk will download the result and will save it as a dataframe.

```{r}
results <- 
  bq_project_query(project_id, my_sql_query)
mesalamine_products <- 
  bq_table_download(results)
```

## Visualization

Now let's use tidyverse and look at the number of different dosage forms at the companies (top 5):

```{r}
dosage_counts <- mesalamine_products %>% 
  group_by(applicant_full_name) %>% 
  summarize(num_of_dosage_form = n_distinct(dosage_form)) %>% 
  arrange(desc(num_of_dosage_form)) %>% 
  head(5)
```

Within r, it takes only a few lines of code to visualize the result.

```{r}
ggplot(dosage_counts, aes(x = reorder(applicant_full_name, -num_of_dosage_form), y = num_of_dosage_form)) +
  geom_col() +
  labs(
    title = "Top 5 Pharma Companies by Mesalamine Dosage Form Diversity",
    x = NULL,
    y = "Number of Distinct Dosage Forms"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    plot.title = element_text(size = 14)
  )
```

## Summary

According to the 2016 Chembl Database on Google Cloud, Mylan and Zydus Pharmaceuticals marketed the most mesalamin dosage forms (3 each).

## Disclaimer

This project was created with guidance from a YT Webinar from the R Consortium, titled **Google BigQuery: First Steps in R** [(link)](https://youtu.be/VzteK2K0gbc?si=aKipfdAklNyN-s_e) presented by Joy Payton.

04 June 2025, Huntingdon UK
